---
title: "4. Tutorial: Working with Human Protein Atlas (HPA) xml files offline"
author:
- name: Anh N. Tran
  affiliation: Northwestern University, Illinois, USA
  email: trannhatanh89@gmail.com
date: 6/7/2019
output: 
    BiocStyle::html_document:
        toc: true
        toc_depth: 2
        toc_float: true
        number_sections: true
vignette: >
  %\VignetteIndexEntry{"4. Tutorial: Working with Human Protein Atlas (HPA) xml files offline"}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse=TRUE,
  comment="#>",
  warning=FALSE,
  error=FALSE,
  eval=FALSE
)
```

```{r library, message=FALSE, warning=FALSE, error=FALSE}
library(BiocStyle)
library(HPAanalyze)
library(dplyr)
library(xml2)
```

# The case 

The Human Protein Atlas allow you to download very detailed data for each protein in the form of an xml file, and `hpaXmlGet` and `hpaXml` allow you to retrieve those files automatically from HPA server and parse them. However, due to technical limitation, you will not be able to save those `"xml_document"/"xml_node"` objects. The question is: How do you keep a version of these files to use when you are not connected to the internet, or for reproducibility?

# The solution

## Download and keep a local version of the xml files for yourself

Look at the ["Downloadable data"](https://www.proteinatlas.org/about/download) page from HPA website, you will see how these files are downloaded. Basically, you add `[ensembl_id].xml` to `http://www.proteinatlas.org` to download individual entries (that's what `hpaXmlGet` does behind the scene), or download the [whole big set](https://www.proteinatlas.org/download/proteinatlas.xml.gz).

From there, you can import the file using `xml2::read_xml()`. The output should be exactly the same as `hpaXmlGet`.

```{r}
## same as hpaXmlGet("ENSG00000134057")
CCNB1xml <- xml2::read_xml("data/ENSG00000134057.xml")
```

## Business as usual with `hpaXml` functions

Since the umbrella function `hpaXml` take either the *ensembl id* or the imported `xml_document` object, you can feed what you just imported to it and get the expected result.

```{r}
CCNB1_parsed <- hpaXml(CCNB1xml)
```

You can obviously use other `hpaXml` functions as well.

```{r}
hpaXmlProtClass(CCNB1xml)
hpaXmlTissueExprSum(CCNB1xml)
hpaXmlAntibody(CCNB1xml)
hpaXmlTissueExpr(CCNB1xml)
```

## Save your parsed objects

It is recommended that you save your parsed objects for reproducibility. Unlike the `xml_document` object, these parsed objects are just regular R lists of standard vectors or data frames. You can save them just as usual.

```{r}
saveRDS(CCNB1_parsed, "data/CCNB1_parsed.rds")
```

# Copyright

__Anh Tran, 2018-2022 __

Please cite: __Tran, A.N., Dussaq, A.M., Kennell, T. et al. HPAanalyze: an R package that facilitates the retrieval and analysis of the Human Protein Atlas data. BMC Bioinformatics 20, 463 (2019) https://doi.org/10.1186/s12859-019-3059-z __
