---
title: "Working with HPAanalyze package"
author:
- name: Anh N. Tran
  affiliation: University of Alabama at Birmingham, Alabama, USA
  email: trannhatanh89@gmail.com
- name: Anita B. Hjelmeland
  affiliation: University of Alabama at Birmingham, Alabama, USA
date: "`r Sys.Date()`"
output: 
    BiocStyle::html_document:
        toc: true
        toc_depth: 2
        toc_float: true
        number_sections: true
vignette: >
  %\VignetteIndexEntry{"Working with HPAanalyze package""}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  error = FALSE
)

library(BiocStyle)
library(HPAanalyze)
library(tibble)
library(dplyr)

BiocStyle::markdown(css.files = c('styles.css'))
```

# Summary

* **Background:** The Human Protein Atlas is...
* **Results:** `HPAanalyze` is an R package for retreiving and performing exploratory data analysis from HPA. It provides functionality for importing data tables and xml files from HPA, exporting and visualizing data, as well as download all staining images of interest. The package is free, open source, and available via Github.
* **Conclusions:** `HPAanalyze` intergrates into the R workflow via the `tidyverse` philosophy and data structures, and can be used in combination with Bioconductor packages for easy analysis of HPA data.
  
* **Keywords:** Human Protein Atlas, Proteomics, Homo Sapiens, Visualization, Software

# Background

HPAanalyze consists of

# Implementation

*HPAanalyze* is and R package with a MIT licence and is designed for easy retreiving and performing exploratory data analysis from HPA. It provides functionality for importing data tables and xml files from HPA, exporting and visualizing data, as well as download all staining images of interest. 

## The different HPA data formats

The Human Protein Atlas project provides data via two main mechanisms: *Full datasets* in the form of downloadable compressed tab-separated files (.tsv) and *individual entries* in XML, RDF and TSV formats. The full downloadable datasets includes normal tissue, pathology (cancer), subcellular location, RNA gene and RNA isoform data. For individual entries, the XML format is the most comprehensive, providing information on the target protein, antibodies, summary for each tissue and detailed data from each sample including clinical data, IHC scoring and image download links.

## *HPAanalyze* overview

*HPAanalyze* is designed to fullfill 3 main tasks: (1) Import, subsetting and export downloadable datasets; (2) Visualization of downloadable datasets for exploratory analysis; and (3) Working with the individual XML files. This package aims to serve researchers with little programming experience, but also allow power users to use the imported data as desired.

### Obtaining *HPAanalyze*

The development version of *HPAanalyze* is available on Github can be installed with:  
  
`devtools::install_github("trannhatanh89/HPAanalyze")`

### Full dataset import, subsetting and export

The `hpaDownload()` function downloads full datasets from HPA and imports them into R as a list of tibbles, the standard object of *tidyverse*, which can subsequently be subset with `hpaSubset()` and export into .xmlx files with `hpaExport()`. The standard object allow the imported data to be further processed in a traditional R workflow. The ability to quickly subset and export data gives researchers the option to use other non-R downstream tools, such as GraphPad for creating publication-quality graphics, or share a subset of data containing only proteins of interest.

### Visualization

The `hpaVis` function family take the output of `hpaDownload()` (or `hpaSubset()`) provides quick visualization of the data, with the intention of aiding exploratory analysis. Nevertheless, the standard `ggplot` object output of these functions give users the option to further customize the plots for publication. All `hpaVis` functions share the same syntax for arguments: subsetting, specifying colors and opting to use custom themes.

The first release of the *HPAanalyze* package includes three functions: `hpaVisTissue()` for the *normal tissue*, `hpaVisPatho()` for the *pathology/cancer*, and `hpaVisSubcell()` for the *subcellular location* datasets.

### Individual xml import and image downloading

The `hpaXml` function family import and extract data from individual XML entries from HPA. The `hpaXmlGet()` function downloads and imports data as  "xml_document"/"xml_node" object, which can subsequently be processed by other `hpaXml` functions. The XML format from HPA contains a wealth of information that may not be covered by this package. However, users can extract any data of interest from the imported XML file using the xml2 package.

In the first release, *HPAanalyze* includes four functions for data extraction from HPA XML files: `hpaXmlProtClass()` for protein class information, `hpaTissueExprSum()` for summary of protein expression in tissue, `hpaXmlAntibody()` for a list of antibody used to stain for the protein of interest, and `hpaTissueExpr()` for a detailed data from each sample including clinical data and IHC scoring.

`hpaTissueExprSum` and `hpaTissueExpr` provide download links to download relevant staining images, with the former function also gives the options to automate the downloading process.

### Compatibility with `hpar` Bioconductor package

Functionality| hpar                            | HPAanalyze 
------------:|:--------------------------------|:--------------------------------
Datasets     | Included in package             | Download from server or Import from hpar
Query        | Ensembl id                      | HCNC symbol for datasets, Ensembl id for XML
Data version | One stable version              | Latest by default, option to download older
Release info | Access via functions            | N/A
View relevant browser page | Via `getHPA` function | N/A
Visualization| N/A                             | Exploratory via `hpaVis` functions
XML          | N/A                             | Download and import via `hpaXml` functions
Histology image| View by loading browser page  | Extract links via `hpaXml` functions

: (\#tab:table) Complementary functionality between hpar and HPAanalyze

## Working with HPA downloadable datasets

A typical workflow with HPA downloadable datasets using *HPAanalyze* consist of the following steps: (1) Download and import data into R with `hpaDownload()`; (2) View available parameter for subseting with `hpaListParam()`; (3) Subset data with `hpaSubset()`; and (4) (optional) Export data with `hpaExport`.

### Download and import data with `hpaDownload()`

```{r downloaded_data, eval=FALSE}
downloaded_data <- hpaDownload(downloadList = 'all')
summary(downloaded_data)

#>                          Length Class  Mode
#> normal_tissue             6     tbl_df list
#> pathology                11     tbl_df list
#> subcellular_location     11     tbl_df list
#> rna_tissue                5     tbl_df list
#> rna_cell_line             5     tbl_df list
#> transcript_rna_tissue     4     tbl_df list
#> transcript_rna_cell_line  4     tbl_df list
```

The `normal_tissue` dataset contains information about protein expression profiles in human tissues based on IHC staining. The datasets contain six columns: `ensembl` (Ensembl gene identifier); `gene` (HGNC symbol), `tissue` (tissue name); `cell_type` (annotated cell type); `level` (expression value); `reliability` (the gene reliability of the expression value).

```{r normal_tissue}
downloaded_data <- hpaDownload(downloadList = 'histology', version = 'example')
tibble::glimpse(downloaded_data$normal_tissue, give.attr = FALSE)
```

The `pathology` dataset contains information about protein expression profiles in human tumor tissue based on IHC staining. The datasets contain eleven columns: `ensembl` (Ensembl gene identifier); `gene` (HGNC symbol); `cancer` (cancer type); `high`, `medium`, `low`, `not_detected` (number of patients annotated for different staining levels); `prognostic_favorable`, `unprognostic_favorable`, `prognostic_unfavorable`, `unprognostic_unfavorable` (log-rank p values for patient survival and mRNA correlation).

```{r pathology}
tibble::glimpse(downloaded_data$pathology, give.attr = FALSE)
```

The `subcellular_location` dataset contains information about subcellular localization of proteins based on IF stanings of normal cells. The datasets contain eleven columns: `ensembl` (Ensembl gene identifier); `gene` (HGNC symbol); `reliability` (gene reliability score); `enhanced` (enhanced locations); `supported` (supported locations); `approved` (approved locations); `uncertain` (uncertain locations); `single_cell_var_intensity` (locations with single-cell variation in intensity); `single_cell_var_spatial` (locations with spatial single-cell variation); `cell_cycle_dependency` (locations with observed cell cycle dependency); `go_id` (Gene Ontology Cellular Component term identifier).

```{r subcellular_location}
tibble::glimpse(downloaded_data$subcellular_location, give.attr = FALSE)
```

The `rna_tissue` and `rna_cell_line` datasets contain RNA expression levels of 37 tissues and 64 cell lines based on RNA-seq. These datasets contain four columns each: `ensembl` (Ensembl gene identifier); `gene` (HGNC symbol); `tissue`/`cell_line` (type of sample); `value` + `unit` (expression level measured by transcripts per million).

```{r rna, warning=FALSE, message=FALSE}
downloaded_data <- hpaDownload(downloadList = 'rna', version = 'hpar')

tibble::glimpse(downloaded_data$rna_tissue, give.attr = FALSE)

tibble::glimpse(downloaded_data$rna_cell_line, give.attr = FALSE)
```

Similarly, the `transcript_rna_tissue` and `transcript_rna_cell_line` datasets contain RNA isoform levels. These datasets contain four columns each: `ensembl` (Ensembl gene identifier); `transcript` (Ensembl transcript identifier); `tissue`/`cell_line` (type of sample); `value` (expression level measured by transcripts per million). Note that these datasets are significantly larger than others and should only be downloaded when necessary.

```{r eval=FALSE}
downloaded_data <- hpaDownload(downloadList = 'isoform', version = 'v18')

tibble::glimpse(downloaded_data$transcript_rna_tissue, give.attr = FALSE)

#> Observations: 27,535,996
#> Variables: 4
#> $ ensembl    <chr> "ENSG00000000003", "ENSG00000000003", "ENSG0000000...
#> $ transcript <chr> "ENST00000373020", "ENST00000494424", "ENST0000049...
#> $ tissue     <chr> "adipose tissue.V1", "adipose tissue.V1", "adipose...
#> $ value      <dbl> 27.3577003, 0.0000000, 1.9341500, 1.6059300, 0.000...

tibble::glimpse(downloaded_data$transcript_rna_cell_line, give.attr = FALSE)

#> Observations: 20,972,183
#> Variables: 4
#> $ ensembl    <chr> "ENSG00000000003", "ENSG00000000003", "ENSG0000000...
#> $ transcript <chr> "ENST00000373020", "ENST00000494424", "ENST0000049...
#> $ cell_line  <chr> "A-431.C35", "A-431.C35", "A-431.C35", "A-431.C35"...
#> $ value      <dbl> 29.406799, 0.000000, 0.992916, 0.398387, 0.239204,...
```

Optionally, data can be imported from the *hpar* package. Please note that the *hpar* package does not contain the RNA isoform datasets, which are very large and should be downloaded only when necessary.

For data from *hpar*, release information can be accessed with `getHpaVersion`, `getHpaDate` and `getHpaEnsembl`.

```{r hpar, eval=-1}
hpaDownload('all', 'hpar')

hpar::getHpaVersion()

hpar::getHpaDate()

hpar::getHpaEnsembl()
```


### List available data with `hpaListParam()`

To see what parameters are available for subsequent subsetting/visualizing, HPAanalyze includes the function `hpaListParam()`. The input for this function is the output of `hpaDownload`.

```{r list_param, eval=FALSE}
downloaded_data <- hpaDownload(downloadList = 'all')
str(hpaListParam(downloaded_data))

#> List of 6
#>  $ normal_tissue       : chr [1:58] "adrenal gland" "appendix" "bone marrow" "breast" ...
#>  $ normal_cell         : chr [1:82] "glandular cells" "lymphoid tissue" "hematopoietic cells" "adipocytes" ...
#>  $ cancer              : chr [1:20] "breast cancer" "carcinoid" "cervical cancer" "colorectal cancer" ...
#>  $ subcellular_location: chr [1:32] "Cytosol" "Mitochondria" "Aggresome" "Plasma membrane" ...
#>  $ normal_tissue_rna   : chr [1:37] "adipose tissue" "adrenal gland" "appendix" "bone marrow" ...
#>  $ cell_line_rna       : chr [1:64] "A-431" "A549" "AF22" "AN3-CA" ...
```

### Subset data with `hpaSubset()`

`hpaSubset()` filters the output of `hpaDownload()` for desirable target genes, tissues, cell types, cancer, and cell lines. The data will be subset only where applicable (i.e. `normal_tissue` will not be subset by *cancer*). The main purpose of `hpaSubset` is to prepare a manageable set of data to be exported. However, this function may also be useful for other data table manipulation purposes.

```{r subset, message=FALSE, warning=FALSE}
downloaded_data <- hpaDownload(downloadList = 'histology', version = 'example')
sapply(downloaded_data, nrow)

gene_list <- c('TP53', 'EGFR', 'CD44', 'PTEN', 'IDH1', 'IDH2', 'CYCS')
tissue_list <- c('breast', 'cerebellum', 'skin 1')
cancer_list <- c('breast cancer', 'glioma', 'melanoma')
cell_line_list <- c('A-431', 'A549', 'AF22', 'AN3-CA')

subset_data <- hpaSubset(data = downloaded_data,
                         targetGene = gene_list,
                         targetTissue = tissue_list,
                         targetCancer = cancer_list,
                         targetCellLine = cell_line_list)
sapply(subset_data, nrow)
```

### Export data with `hpaExport()`
As the name suggests, `hpaExport()` exports the output of `hpaSubset()` to one .xlsx file. Each dataset is placed in a separate sheet. More formats such as .csv and .tsv might be added in future release; hence, the `fileType` argument is included. 

```{r eval=FALSE}
hpaExport(subset_data, fileName = 'subset.xlsx', fileType = 'xlsx')
```

## Visualize downloaded data

`HPAanalyze` provides the ability to quickly visualize data from downloaded HPA datasets with the `hpaVis` function family. The goal of these functions is to aid exploratory analysis of a group of target genes, which maybe particularly useful for gaining insights into pathways or gene signatures of interest.

The `hpaVis` functions share a common syntax, where the input (`data` argument) is the output of `hpaDownload()` or `hpaSubset()` (although they do their own subseting so it is not necessary to use `hpaSubset()` unless you want to reduce the size of your data object). Depending on the function, the `target` arguments will let you choose to visualize your vectors of genes, tissue, cell types, etc. (See the help files for more details.) All of `hpaVis` functions generate standard `ggplot2` plots, which allow you to further customize colors and themes. Colors maybe changed via the `color` argument, while the default theme maybe overriden by setting the `customTheme` argument to `FALSE`.

Currently, the `normal_tissue`, `pathology` and `subcellular_location` data can be visualized, with more functions planned for future releases.

```{r echo=FALSE, warning=FALSE, message=FALSE}
downloaded_data <- hpaDownload('histology', 'example')
```


### Visualize tissue data with `hpaVisTissue()`

`hpaVisTissue()` generates a "heatmap", in which the expression of proteins of interest (quantified IHC staining) are plotted for each cell type of each tissue. 

```{r fig.width=6}
gene_list <- c('TP53', 'EGFR', 'CD44', 'PTEN', 'IDH1', 'IDH2', 'CYCS')
tissue_list <- c('breast', 'cerebellum', 'skin 1')

hpaVisTissue(downloaded_data,
             targetGene = gene_list,
             targetTissue = tissue_list)
```


### Visualize expression in cancer with `hpaVisPatho()`

`hpaVisPatho()` generates an arrays of column graphs showing the expression of proteins of interest in each cancer. 

This example also demonstrate how the colors of the graphs could be customized, which is a common functionality of the `hpaVis` family.

```{r fig.width=6}
gene_list <- c('TP53', 'EGFR', 'CD44', 'PTEN', 'IDH1', 'IDH2', 'CYCS')
cancer_list <- c('breast cancer', 'glioma', 'lymphoma', 'prostate cancer')
color_gray <- c('slategray1', 'slategray2', 'slategray3', 'slategray4')

hpaVisPatho(downloaded_data,
            targetGene = gene_list,
            targetCancer = cancer_list,
            color = color_gray)
```


### Visualize subcellular location data with `hpaVisSubcell()`

`hpaVisSubcell()` 

```{r fig.width=6}
gene_list <- c('TP53', 'EGFR', 'CD44', 'PTEN', 'IDH1', 'IDH2', 'CYCS')

hpaVisSubcell(downloaded_data,
              targetGene = gene_list)
```


## Working with individual xml files

### Import xml file with `hpaXmlGet()`

```{r}
CCNB1_xml <- hpaXmlGet('ENSG00000134057')
```

### View protein classes with `hpaXmlProtClass()`

```{r}
hpaXmlProtClass(CCNB1_xml)
```

### Get summary and images of tissue expression with `hpaXmlTissueExprSum()`

```{r}
hpaXmlTissueExprSum(CCNB1_xml)
```

### Get details of individual IHC samples with `hpaXmlAntibody()` and `hpaXmlTissueExpr()`

```{r}
hpaXmlAntibody(CCNB1_xml)

tissue_expression <- hpaXmlTissueExpr(CCNB1_xml)

summary(tissue_expression)

tissue_expression[[1]]

```

## Future developments

# Conclusion

# Availability and requirements

* Project name: HPAanalysis
* Project home page: https://github.com/trannhatanh89/HPAanalyze
* Operating system(s): All platforms where R is available, including Windows, Linux, OS X
* Programming language: R
* Other requirements: R 3.5.0 or higher, and the R packages dplyr,     XLConnect, ggplot2, readr, tibble, xml2, reshape2, tidyr, stats, utils, and hpar
* License: MIT
* Any restrictions to use by non-academics: Freely available to everyone

# Abbreviations

HPA: The Human Protein Atlas; 

# Declarations

## Acknowledgements

I appreciate the support of the National institutes of Health National Cancer Institute R01 CA151522 and funds from the Department of Cell, Developmental and Integrative Biology at the University of Alabama at Birmingham.

## Authors' contributions

ANT created the R package and drafted the manuscript. ABH supervised the project and revised the manuscript. All authors read and approved the final manuscript.

# References

# Copyright
© Anh Tran et al. 2018

# Session info
```{r echo=FALSE}
sessionInfo()
```

