---
title: "HPAanalyze quick start guide"
author:
- name: Anh N. Tran
  affiliation: University of Alabama at Birmingham, Alabama, USA
  email: trannhatanh89@gmail.com
- name: Anita B. Hjelmeland
  affiliation: University of Alabama at Birmingham, Alabama, USA
date: "`r Sys.Date()`"
output: 
    BiocStyle::html_document:
        toc: true
        toc_depth: 2
        toc_float: true
        number_sections: true
vignette: >
  %\VignetteIndexEntry{"Working with HPAanalyze package"}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse=TRUE,
  comment="#>",
  warning=FALSE,
  error=FALSE
)
```

```{r library}
library(BiocStyle)
library(HPAanalyze)
library(tibble)
library(dplyr)
library(ggplot2)
library(hpar)
```


# Visualize protein expression data

Currently, this is available for the normal tissue, pathology (cancers) and subcellular location datasets. The fastest and easiest way is to use the defaults of `hpaVis`. Of course, we cannot visualize everything in those big datasets, so some defauts will be used and you will receive some warning messages.

```{r hpaVis_eg}
hpaVis()
```

```{r}
functionx <- function(x = 1, ...) {
    ls()
}
    
functionx(y = 1)
```


# Copyright
Anh Tran, 2018

# Session info
```{r echo=FALSE}
sessionInfo()
```

