---
title: "HPAanalyze quick start guide"
author:
- name: Anh N. Tran
  affiliation: University of Alabama at Birmingham, Alabama, USA
  email: trannhatanh89@gmail.com
- name: Anita B. Hjelmeland
  affiliation: University of Alabama at Birmingham, Alabama, USA
date: "`r Sys.Date()`"
output: 
    BiocStyle::html_document:
        toc: true
        toc_depth: 2
        toc_float: true
        number_sections: true
vignette: >
  %\VignetteIndexEntry{"HPAanalyze quick start guide: Acquire and visualize the Human Protein Atlas data"}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse=TRUE,
  comment="#>",
  warning=FALSE,
  error=FALSE
)
```

```{r library, message=FALSE, warning=FALSE, error=FALSE}
library(BiocStyle)
library(HPAanalyze)
library(dplyr)
```


# Visualize protein expression data

Currently, this is available for the normal tissue, pathology (cancers) and subcellular location datasets. The fastest and easiest way is to use the defaults of `hpaVis`.

```{r hpaVis_eg}
hpaVis(targetGene = c("GCH1", "PTS", "SPR", "DHFR"),
       targetTissue = c("cerebellum", "cerebral cortex", "hippocampus"),
       targetCancer = c("glioma"))
```

Of course, we cannot visualize everything in those big datasets, so some defauts will be used and you will receive some warning messages.

```{r hpaVis_eg2, eval = FALSE}
hpaVis()

# No data provided. Use version 18.
# targetGene variable not specified, default to TP53, RB1, MYC, KRAS and EGFR.
# targetTissue variable not specified, default to breast.
# targetCellType variable not specified, visualize all.
# targetCancer variable not specified, default to breast cancer
# Use hpaListParam() to list possible values for target variables.

```

You can also use `hpaVis` to show just one or two of the three graphs.

```{r hpaVis_eg3}
hpaVis(visType = "Patho",
       targetGene = c("GCH1", "PTS", "SPR", "DHFR"),
       targetCancer = c("glioma", "breast cancer"))
```

One exception though, if you want to plot all cancers, use `hpaVisPatho` with `targetCancer = NULL` (default).

```{r hpaVisPatho_eg}
hpaVisPatho(targetGene = c("GCH1", "PTS", "SPR", "DHFR"))
```

There are many ways you could customize your plots. Please see the documentation for more details.

```{r doc_ex1, eval = FALSE}
?hpaVis # the easy umbrella to visualize protein expression levels
?hpaVisTissue # in normal tissue
?hpaVisSubcell # in subcellular compartments
?hpaVisPatho # in cancers
```

If you want to know what kind of data you can visualize, use `hpaListParam`. You will receive a list of parameters you can use. Please note that if you ask the functions to plot anything that is not on this list, they will just ignore it and plot what are available.

```{r listParam_ex, eval = FALSE}
hpaListParam()
```
```{r listParam_ex2, echo= FALSE}
 hpaListParam() %>% glimpse()
```


# Copyright
Anh Tran, 2018

# Session info
```{r echo=FALSE}
sessionInfo()
```

